---
title: "Patriotism and the Choice Process: National Pride After 9/11"
author: "Omar Lizardo"
date: today
format:
  html:
    toc: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

Which groups are more likely to experience increased feelings of national pride after the experience of a threat to nation? In this paper, I use survey data on a representative sample of Americans collected before and after the events of 9/11 to shed light on this question. This "natural experiment" allows me adjudicate between two competing hypotheses regarding the socio-demographic distribution of feeling of national pride after the experience of a threat to the nation.

Theories in the tradition of the "authoritarian personality" conceive of national pride as reflecting a form "irrational conservatism." These theories predict that members of low status groups will be much more likely to experience renewed feelings of national pride after the experience of a threat to the nation. From this perspective, a threat to the nation is seen as endangering the "ontological security" (Giddens 1984) of persons whose lives are characterized by routine and alienating work and who are disengaged from established institutions.

Theories of attachment to collectivities as a nested group on the other hand (Lawler 1992; Mueller and Lawler 1999), point to a different dynamic, and lead to a clearly divergent empirical implication. According to the choice-process approach, individuals are more likely to experience affective attachment to a larger group within which other groups that they belong to is nested (such as the nation), only to the extent that they are able to in fact experience freedom and autonomy in their everyday dealings, and to the extent that this autonomy and freedom is attributed to membership both in the more proximate groups that they are part of, and to the larger unit within which these groups are nested (Lawler 1992).

# Data and Methods

In this modernized workflow, we draw data directly from the General Social Survey (GSS) using reproducible queries, focusing on the years 1996 and 2004 (before and after the events of 9/11). We use the `gssr` package to extract data and handle sample weights appropriately.

```{r}
#| label: setup-and-data
#| include: false

library(gssr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(broom)
library(modelsummary)
library(haven)
library(marginaleffects)
library(psych)

# Fetch data for 1996 and 2004
gss96 <- gss_get_yr(1996)
gss04 <- gss_get_yr(2004)

# Relevant variables
vars_to_keep <- c("year", "age", "educ", "degree", "sex", "race", "polviews", "born", 
                  "reg16", "region", "sei", "wtssall", "belikeus", "ifwrong")

# Combine datasets, only keeping available variables across both datasets to avoid errors
gss_comb <- bind_rows(gss96, gss04) %>%
  select(any_of(vars_to_keep), 
         starts_with("proud"), 
         starts_with("am"), 
         starts_with("mem")) %>%
  filter(year %in% c(1996, 2004))
```

We process the U.S. Pride scale and Nationalist scale mirroring the original Stata script, ensuring proper handling of missing values ("IAP", "DK", "NA") and dichotomization. We also construct predictors such as cohort, education categories, and voluntary association memberships.

```{r}
#| label: data-processing
#| echo: true

gss_clean <- gss_comb %>%
  # Strip labelling attributes for simpler manipulation
  mutate(across(everything(), as.numeric)) %>%
  
  # For proud variables: Stata dropped if == 0
  filter(if_all(prouddem:proudgrp, ~ .x != 0 | is.na(.x))) %>%
  
  mutate(
    # Recodes 0 and 9 to NA for am* variables
    across(ambornin:amsports, ~ na_if(.x, 0)),
    across(ambornin:amsports, ~ na_if(.x, 9)),
    
    # Dichotomize proud* variables (1 = yes)
    across(prouddem:proudgrp, ~ ifelse(.x == 1, 1, 0), .names = "{.col}bin"),
    
    # Dichotomize am* variables for patriotism scale
    across(ambornin:amfeel, ~ ifelse(.x == 1, 1, 0), .names = "{.col}bin"),
    
    # Recode amcitizn-ifwrong: 8 = 3, then dichotomize < 3
    across(c(amcitizn, amshamed, belikeus, ambetter, ifwrong), ~ ifelse(.x == 8, 3, .x)),
    across(c(amcitizn, amshamed, belikeus, ambetter, ifwrong), ~ ifelse(.x < 3, 1, 0), .names = "{.col}bin"),
    
    # Demographic recodes
    educ = ifelse(educ %in% c(98, 99), NA, educ),
    degree = ifelse(degree %in% c(8, 9), NA, degree),
    memnum = ifelse(memnum == -1, NA, ifelse(memnum %in% c(98, 99), 0, memnum)),
    
    usaborn = ifelse(born == 1, 1, 0),
    south = ifelse(reg16 > 4 & region < 8, 1, 0),
    conservative = ifelse(polviews %in% c(6, 7), 1, 0),
    
    # Membership categories and binaries
    memnumcat = case_when(
      memnum == 0 ~ "0",
      memnum %in% 1:2 ~ "1-2",
      memnum >= 3 ~ "3+",
      TRUE ~ NA_character_
    ),
    memnumcat = factor(memnumcat, levels = c("0", "1-2", "3+")),
    
    educcat = case_when(
      educ <= 6 ~ "0-6 years",
      educ >= 7 & educ <= 11 ~ "7-11 years",
      educ == 12 ~ "12 years",
      educ >= 13 & educ <= 15 ~ "13-15 years",
      educ == 16 ~ "16 years",
      educ >= 17 & educ <= 19 ~ "17-19 years",
      educ >= 20 ~ "20+ years",
      TRUE ~ NA_character_
    ),
    educcat = factor(educcat, levels = c("0-6 years", "7-11 years", "12 years", 
                                         "13-15 years", "16 years", "17-19 years", "20+ years")),
    
    cohort = year - age
  ) %>%
  # Dichotomize the individual membership types
  mutate(across(memfrat:memchurh, ~ ifelse(.x > 0 & !is.na(.x) & .x == 1, 1, 0), .names = "{.col}bin")) %>%
  rowwise() %>%
  mutate(
    # U.S. Pride scale: sum of binary proud items. NA if any are missing.
    uspridescal = if(any(is.na(c_across(prouddembin:proudgrpbin)))) NA_real_ 
                  else sum(c_across(prouddembin:proudgrpbin)),
    
    # Patriotism scale
    patriotscal = if(any(is.na(c_across(c(amborninbin, amcitbin, amlivedbin, amgovtbin, amfeelbin))))) NA_real_ 
                  else sum(c_across(c(amborninbin, amcitbin, amlivedbin, amgovtbin, amfeelbin))),
                  
    # Nationalist scale (mean reversed)
    nationalistscal = 6 - mean(c_across(c(amcitizn, belikeus, ambetter, ifwrong)), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  # Filter missing data out of the weight to ensure smooth modeling
  filter(!is.na(wtssall))

# Ensure appropriate factors for models
gss_clean <- gss_clean %>%
  mutate(
    year_f = factor(year),
    sex_f = factor(sex),
    race_f = factor(race)
  )
```

# Results

To make the regression tables cleaner, we first define a dictionary of clean variable labels.

```{r}
#| label: setup-labels
#| echo: false

coef_labels <- c(
  "educ" = "Education (Years)",
  "cohort" = "Birth Cohort",
  "year_f2004" = "Year: 2004",
  "educ:year_f2004" = "Education × Year 2004",
  "sei" = "Occupational Status (SEI)",
  "sex_f2" = "Female",
  "race_f2" = "Race: Black",
  "race_f3" = "Race: Other",
  "south" = "Region: South",
  "nationalistscal" = "Nationalist Scale",
  "memnum" = "Total Memberships (Count)",
  "memnumcat1-2" = "Memberships: 1-2",
  "memnumcat3+" = "Memberships: 3+",
  "memfratbin" = "Fraternal Group",
  "memservbin" = "Service Club",
  "memvetbin" = "Veterans Group",
  "mempolitbin" = "Political Club",
  "memunionbin" = "Labor Union",
  "memsportbin" = "Sports Group",
  "memyouthbin" = "Youth Group",
  "memschlbin" = "School Service",
  "memhobbybin" = "Hobby Club",
  "memgreekbin" = "School Fraternity",
  "memnatbin" = "Nationality Group",
  "memfarmbin" = "Farm Organization",
  "memlitbin" = "Literary/Art Group",
  "memprofbin" = "Professional Society",
  "memchurhbin" = "Church Group",
  "(Intercept)" = "Intercept"
)
```

We first visualize the change in the U.S. Pride Scale score between 1996 and 2004. As shown in @fig-pride-boxplot, national pride distributions shifted post-9/11.

```{r}
#| label: fig-pride-boxplot
#| fig-cap: "U.S. Pride Scale Score by Year"

ggplot(gss_clean %>% filter(!is.na(uspridescal)), 
       aes(x = factor(year), y = uspridescal, weight = wtssall)) +
  geom_boxplot(fill = "lightblue") +
  labs(x = "Year", y = "U.S. Pride Scale Score") +
  theme_minimal()
```

## Scale Validation

Before running models, we assess the internal reliability of the scales constructed for this analysis.

```{r}
#| label: scale-reliability
#| echo: false

# Calculate Cronbach's alpha for the U.S. Pride scale
pride_items <- gss_clean %>% select(prouddembin:proudgrpbin)
pride_alpha <- psych::alpha(pride_items, check.keys = FALSE)$total$raw_alpha

# Calculate Cronbach's alpha for the Patriotism scale
patriot_items <- gss_clean %>% select(amborninbin, amcitbin, amlivedbin, amgovtbin, amfeelbin)
patriot_alpha <- psych::alpha(patriot_items, check.keys = FALSE)$total$raw_alpha

# Calculate Cronbach's alpha for the Nationalism scale
nat_items <- gss_clean %>% select(amcitizn, belikeus, ambetter, ifwrong)
nat_alpha <- psych::alpha(nat_items, check.keys = FALSE)$total$raw_alpha

cat("Cronbach's alpha for U.S. Pride scale:", round(pride_alpha, 3), "\n")
cat("Cronbach's alpha for Patriotism scale:", round(patriot_alpha, 3), "\n")
cat("Cronbach's alpha for Nationalism scale:", round(nat_alpha, 3), "\n")
```

## Binary Item Models

We examine how the individual likelihoods of being "very proud" of various aspects of the U.S. changed from 1996 to 2004 by running survey-weighted logistic regressions for each binary indicator.

```{r}
#| label: tbl-logistic-pride
#| results: asis

# Identify binary pride variables
pride_bins <- grep("^proud.*bin$", names(gss_clean), value = TRUE)

logit_models <- list()
for(v in pride_bins) {
  # formula: var ~ year_f + cohort
  f <- as.formula(paste(v, "~ year_f + cohort"))
  m <- glm(f, data = gss_clean, family = quasibinomial(), weights = wtssall)
  logit_models[[v]] <- m
}

# Show a selection of these models, converting estimates to Odds Ratios
modelsummary(logit_models[1:5], 
             stars = TRUE, 
             exponentiate = TRUE,
             coef_map = coef_labels,
             align = "lccccc",
             title = "Logistic Regressions of Individual Pride Items (Odds Ratios)",
             output = "markdown")
```

## Nationalism and Irrational Conservatism

Before analyzing U.S. Pride, we can look at the "Nationalism" scale. The "irrational conservatism" perspective posits that 9/11 should have driven an increase in nationalism, particularly among lower-status groups. We estimate whether Nationalism increased overall, and whether the effect of education on Nationalism changed over time.

```{r}
#| label: tbl-nationalism
#| results: asis

# Baseline change in nationalism
mod_nat_base <- lm(nationalistscal ~ year_f + cohort, data = gss_clean, weights = wtssall)

# Interaction of education and period on nationalism
mod_nat_int <- lm(nationalistscal ~ educ * year_f + cohort, data = gss_clean, weights = wtssall)

modelsummary(list("Base Change" = mod_nat_base, "Educ Interaction" = mod_nat_int),
             stars = TRUE,
             coef_map = coef_labels,
             align = "lcc",
             title = "Predictors of Nationalism Scale (1996 vs 2004)",
             output = "markdown")
```

As the Indiana presentation notes indicate, neither the need-based nor the irrational conservatism hypotheses fully explain the patterns of national pride and nationalism after 9/11. We now turn to U.S. Pride under the Choice-Process framework.

## Pride Scale Regression Analysis

We estimate OLS models comparing the effects of education, occupational status (SEI), and nested group memberships on national pride across the two periods. Instead of purely subsetting, we fit a pooled model to formally test the interaction between education and the survey year (1996 vs 2004).

```{r}
#| label: tbl-regression-educ
#| results: asis

# Sub-setted models
mod_educ_96 <- lm(uspridescal ~ educ + cohort, data = gss_clean, subset = year == 1996, weights = wtssall)
mod_educ_04 <- lm(uspridescal ~ educ + cohort, data = gss_clean, subset = year == 2004, weights = wtssall)

# Pooled model with interaction
mod_educ_interaction <- lm(uspridescal ~ educ * year_f + cohort, data = gss_clean, weights = wtssall)

modelsummary(list("1996" = mod_educ_96, "2004" = mod_educ_04, "Pooled Interaction" = mod_educ_interaction), 
             stars = TRUE, 
             coef_map = coef_labels,
             align = "lccc",
             title = "Effects of Education on U.S. Pride Scale (1996 vs 2004)",
             output = "markdown")
```

To better visualize how the effect of education on national pride shifted post-9/11, we plot the predicted marginal effects of education for both years.

```{r}
#| label: fig-marginal-educ
#| fig-cap: "Predicted U.S. Pride Score across Education Levels by Year"
#| echo: false

plot_predictions(mod_educ_interaction, condition = c("educ", "year_f")) +
  labs(x = "Years of Education", y = "Predicted U.S. Pride Score", color = "Year", fill = "Year") +
  theme_minimal()
```

```{r}
#| label: tbl-regression-sei
#| results: asis

mod_sei_96 <- lm(uspridescal ~ sei + cohort, data = gss_clean, subset = year == 1996, weights = wtssall)
mod_sei_04 <- lm(uspridescal ~ sei + cohort, data = gss_clean, subset = year == 2004, weights = wtssall)

modelsummary(list("1996" = mod_sei_96, "2004" = mod_sei_04), 
             stars = TRUE, 
             coef_map = coef_labels,
             align = "lcc",
             title = "Effects of SEI on U.S. Pride Scale (1996 vs 2004)",
             output = "markdown")
```

## Nested Membership and Nationalism (2004)

Finally, following the choice process framework, we assess how voluntary association membership acts as a nested grouping structure that correlates with attachment to the nation. We evaluate this against a full suite of demographic controls (Gender, Race, Region).

```{r}
#| label: tbl-regression-full-2004
#| results: asis

# Baseline with demographics
mod_base <- lm(uspridescal ~ educ + cohort + sex_f + race_f + south, 
               data = gss_clean, subset = year == 2004, weights = wtssall)

# Adding membership dummy (as evaluated in presentation slide 36)
mod_mem_cat <- lm(uspridescal ~ educ + cohort + sex_f + race_f + south + memnumcat, 
                  data = gss_clean, subset = year == 2004, weights = wtssall)

# Adding continuous membership count
mod_mem_cont <- lm(uspridescal ~ educ + cohort + sex_f + race_f + south + memnum, 
                   data = gss_clean, subset = year == 2004, weights = wtssall)

# Full model with Nationalism
mod_full <- lm(uspridescal ~ educ + cohort + sex_f + race_f + south + memnum + nationalistscal, 
               data = gss_clean, subset = year == 2004, weights = wtssall)

modelsummary(list("Demographics" = mod_base, "Mem (Categorical)" = mod_mem_cat, "Mem (Count)" = mod_mem_cont, "Full (+ Nationalism)" = mod_full), 
             stars = TRUE, 
             coef_map = coef_labels,
             align = "lcccc",
             title = "Predictors of U.S. Pride (2004 sample only)",
             output = "markdown")
```

```{r}
#| label: fig-membership-forest
#| fig-cap: "Coefficients of Specific Association Memberships on U.S. Pride (Controlling for Cohort)"
#| echo: false
#| fig-height: 6

# Examining the effects of specific types of organizational memberships
mem_bins <- grep("^mem.*bin$", names(gss_clean), value = TRUE)
mem_bins <- setdiff(mem_bins, c("memnumbin", "memotherbin", "memgrp1bin", "memgrp2bin", "memgrp3bin", "memgrp4bin", "memgrp5bin"))

# Fit models and extract coefficients for each membership type
mem_results <- lapply(mem_bins, function(v) {
  f <- as.formula(paste("uspridescal ~", v, "+ cohort"))
  m <- lm(f, data = gss_clean, weights = wtssall)
  # Extract tidy results and keep only the membership coefficient
  res <- tidy(m, conf.int = TRUE) %>% filter(term == v)
  # Add the clean label for plotting
  res$clean_name <- coef_labels[v]
  res
}) %>% bind_rows()

# Plot the estimates ordered by effect size
ggplot(mem_results, aes(x = estimate, y = reorder(clean_name, estimate), color = clean_name)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0, linewidth = 0.8, alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Effect of Memberships on National Pride",
       x = "Estimate (Change in U.S. Pride Scale)",
       y = "Membership Type") +
  theme_minimal() +
  theme(legend.position = "none")
```

The results above reveal a crucial dynamic at the heart of the choice-process approach. While the overall *volume* of voluntary memberships (the `memnum` count variable in Table 4) does not significantly predict national pride, specific *types* of organizational memberships have strong, distinct effects. A simple count of memberships assumes all groups foster national attachment equally. However, according to choice-process theory, affective attachment to the nation requires that the proximate group be cognitively "nested" within the national structure. Groups that are inherently linked to the national fabric—such as Veterans Groups, Political Clubs, and Fraternal Organizations—foster autonomy that is easily attributed to the larger national collective, resulting in increased U.S. pride. Conversely, when all memberships are aggregated, the positive effects of these highly-nested groups are diluted by the noise of locally-focused or apolitical associations, explaining why sheer volume fails to capture the underlying sociological mechanism.

# Conclusion

By migrating the project to R and Quarto, the data collection, cleaning, and analysis are now fully transparent and reproducible within a single integrated document.
